# Dimensionality reduction techniques

```{r, include=FALSE}
options(knitr.duplicate.label = "allow")
```

## Description

The dataset includes now the following variables:

- gni = Gross National Income per Capita
- leb =  Life Expectancy at Birth 
- eye = Expected Years of Education
- mmr = Maternal Mortality Ratio 
- abr = Adolescent Birth Rate
- prp = Female Percent Representation in Parliament 
- edu.rat = the ratio of Female and Male populations with secondary education (secedu.f/secedu.m)
- lab.rat = the ratio of labour force participation of females and males (lfpr.f/lfpr.m)

## Exploratory analysis

Let's explore the dataset

```{r setup, include=FALSE}
library(GGally)
library(ggplot2)
library(tidyr)
human <- read.csv("/Users/Maria/R/IODS-project/data/human.csv")
str(human)
dim(human)
summary(human)

# attach the row names
row.names(human) <- human[,1]
human <- human[,-1]
str(human)
```



```{r}
# check the distribution
library(corrplot)
ggpairs(human)
cor(human) %>% round(2)
cor(human) %>% corrplot(method="circle", type = "upper", cl.pos = "b", tl.pos = "d", tl.cex = 0.6)
summary(human)
```


From the plot we can the following:

1) the strong positive correlation is between:
- expected years of education **eye** and life expectancy at birth **leb**: **0.79**
- adolescent birth rate **abr** and maternal mortality ratio **mmr**: **0.76**

2) the strong negative correlation is between:
- maternal mortality ratio **mmr** and life expectancy at birth **leb**: **-0.86**
- adolescent birth rate **abr** and life expectancy at birth **leb**: **-0.73**

Only the expected years of education **eye** variable is close to the normal ditribution, other variables are skewed.


## Principal component analysis (PCA)

Let's perform principal component analysis for our dataset
```{r}
# performing principal component analysis (with the SVD method)
pca_human <- prcomp(human)
pca_human
# drawing a biplot of the principal component representation and the original variables
biplot(pca_human, choices = 1:2, cex = c(0.6, 0.6), col = c("brown", "blue"))
```


## Standardization

The result can't be explained, so we need to scale the data.

```{r}
# scaling the variables
human_scaled <- scale(human)
summary(human_scaled)
```

We can see that all the means equal zero.


Then we perform PCA again but on standardized data
```{r}
# perform PCA (SVD method)
human_pca <- prcomp(human_scaled)
human_pca
# drawthe biplot from the PCA result and the original variables
biplot(human_pca, choices = 1:2, cex = c(0.4, 0.6), col = c("brown", "blue"))
```


Now the plot looks much better in terms of explanation. 

We can conclude the following points:

-  maternal mortality ratio **mmr**, adolescent birth rates **abr**, ratio of labour force participation of females and males **lab.rat** contribute to the first principal component (PC1)
- life expectancy at birth **leb**, gross national income **gni**, expected years of education **eye**, female percent representation in parliament **prp**, the ratio of female and male populations with secondary education **edu.rat** contribute to the second principal component (PC2)
- maternal mortality ratio **mmr** and adolescent birth rate **abr** go to one direction, so they present strong positive correlation; however, both variables show strong negative correlation with all other variables 
- life expectancy at birth **leb**, the ratio of female and male populations with secondary education **edu.rat**, gross national income **gni**, expected years of education **eye** have strong positive correlation as far as these four variables go to one direction
- female percent representation in parliament **prp** and ratio of labour force participation of females and males **lab.rat** have strong positive correlation because of the arrows' direction
 

Personal interpretations of the first two principal component dimensions:

- the first principal component shows the high level development of some countries (e.g. Sierra Leone, Burkina Faso, Republic of Chad, etc.) in terms of adolescent birth rates **abr** and maternal mortality ratio **mmr** 
- the second principal component captures the countries that are more developed



## Multiple Correspondence Analysis (MCA)

We download the tea dataset
```{r}
library(FactoMineR)
data(tea)
keep_columns <- c("Tea", "How", "how", "sugar", "where", "lunch")
# new dataset with selected columns
tea_time <- dplyr::select(tea, one_of(keep_columns))
str(tea_time)
dim(tea_time)
```

Then we visualize the dataset:
```{r}
library(ggplot2)
library(tidyr)
gather(tea_time) %>% ggplot(aes(value)) + facet_wrap("key", scales = "free") + geom_bar() + theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8))
```

Now we performe the MCA:
```{r}
mca <- MCA(tea_time, graph = FALSE)
summary(mca)
```


We can see the strongest correlation of "how" and "where" with the first dimension and these variables are also correlated with the second dimension. 


Let's visualize MCA by categories and individuals:

```{r}
# by individuals
plot(mca, invisible=c("ind"), habillage = "quali")
```


```{r}
# by individuals
plot(mca, invisible=c("var"), habillage = "quali")
```


We can see that "how" and "where" are the most similar categories. We can conclude that 

- people who buy tea in chain stores drink tea bags, and 
- people who buy tea either in chain stores or in tea shops drink tea bags and unpackaged tea. - - people who buy tea in tea shops drink unpackaged tea 
- people drink just pure green tea without any additions and as for black and earl grey they add milk and lemon

First dimension explains 15\% of variance, and the second dimension - 14\% of variance.
